---
title: "R starter script"
author: "G.E.Flore"
date: "2025-03-31"
output: html_document
---

# 1.	Use R as a calculator
# add
```{r}
3 + 5
```
# divide
```{r}
12 / 4
```
# raised to the power
```{r}
2^3
```
# Multiply
```{r}
2 * 3
```

# Square root
```{r}
sqrt(16)
```
#2. assign variables
#10 to X
```{r}
x <- 10
```
# 5 to Y
```{r}
y <- 5
```
# Add the two
```{r}
result <- x * y
result
```
# 3.	Create vectors and explore them
# Vector with gene names
```{r}
gene_names <- c("BRCA1", "TP53", "HER2", "EGFR")
```
# Expression for each gene
```{r}
expression <- c(2.5, 1.8, 3.1, 2.0)
```
# Check length
```{r}
length(expression)
```
# Mean
```{r}
mean(expression)
```
#4.	Use data.frame() to make a mini gene expression table
```{r}
df <- data.frame(Gene=gene_names, Expression=expression)
head(df)
```
#Exercise Set 2: First Bioconductor Setup
#1.	Install Bioconductor package manager
```{r}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install()  
```
# 2.	Install and load a basic Bioconductor package
```{r}
BiocManager::install("Biobase")
library(Biobase)
```
#3.	Explore example ExpressionSet object
```{r}
data(sample.ExpressionSet)
sample.ExpressionSet
```
# View expression values
```{r}
exprs(sample.ExpressionSet)[1:5, 1:5]
```
#Exercise Set 3: Plotting Basics
#1.	Barplot gene expressions
```{r}
barplot(expression, names.arg = gene_names,
        main = "Gene Expression",
        ylab = "Expression Level",
        col = "skyblue")
```
#2.	Simple ggplot2 practice
#Install and Load ggplot library
```{r}
install.packages("ggplot2")
library(ggplot2)
```

```{r}
ggplot(df, aes(x = Gene, y = Expression)) +
    geom_col(fill = "tomato") +
    theme_minimal()
```





Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
