---
title: "Filtered scRNAseq _prost"
author: "G.E.Flore"
date: "2025-02-04"
output: html_document
---
# Load libraries
```{r}
library(Seurat)
library(EnhancedVolcano)
library(enrichplot)
library(clusterProfiler)
library(org.Hs.eg.db)
```
# Load data
```{r}
data_raw <- read.table("C:/Users/LENOVO/Desktop/scRNAseq _data/data.matrix.txt", header = TRUE, sep = "\t", row.names = 1)
```
# Create Seurat object
```{r}
filtered_object <- CreateSeuratObject(counts = data_raw)
```
# Calculate percentage of mitochondrila genes
```{r}
filtered_object[["percent.mt"]] <- PercentageFeatureSet(filtered_object, pattern = "^MT-")
```
#Visualize QC metrics
```{r}
VlnPlot(filtered_object, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
```
# Scatter plot to detect outliers
```{r}
FeatureScatter(filtered_object, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
```
# Subset data to keep only high quality counts
```{r}
subset_object <- subset(
  filtered_object,
  subset = nFeature_RNA > 200 & nFeature_RNA < 5000 &
           nCount_RNA > 500 & nCount_RNA < 30000)
```
# Normalize data
```{r}
filtered_object <- NormalizeData(filtered_object, normalization.method = "LogNormalize", scale.factor = 30000)
```
# Identify Highly variable features
```{r}
filtered_object <- FindVariableFeatures(filtered_object, selection.method = "vst", nfeatures = 2000)
```
# Visualize highly variable features
```{r}
VariableFeaturePlot(filtered_object)
```
# Centre and scale the data to prepare for PCA
```{r}
filtered_object <- ScaleData(filtered_object, features = VariableFeatures(object = filtered_object))
```
# Perform PCA to reduce dimentionality with principal components analysis(PCA)
```{r}
filtered_object <- RunPCA(filtered_object, features = VariableFeatures(object = filtered_object))
```
# Visualize PCA results
```{r}
ElbowPlot(filtered_object)
```
# Cluster cells with resolution of 0.5
```{r}
filtered_object <- FindNeighbors(filtered_object, dims = 1:10)
filtered_object <- FindClusters(filtered_object, resolution = 0.5) 
```
# Run UMAP to visuslize cells in @D space
```{r}
filtered_object <- RunUMAP(filtered_object, dims = 1:10)
```
# Visualize UMAP
```{r}
DimPlot(filtered_object, reduction = "umap", label = TRUE)
```
# find marker genes for each cluster to determine their identities
```{r}
cluster_markers <- FindAllMarkers(filtered_object, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)
```
# Express markers for each cluster
```{r}
C0.M <- FindMarkers(filtered_object, ident.1 = 0, min.pct= 0.25, logfc.threshold = 0.25,test.use = "roc", only.pos = TRUE)
C1.M <- FindMarkers(filtered_object, ident.1 = 1, min.pct= 0.25, logfc.threshold = 0.25,test.use = "roc", only.pos = TRUE)
C2.M <- FindMarkers(filtered_object, ident.1 = 2, min.pct= 0.25, logfc.threshold = 0.25,test.use = "roc", only.pos = TRUE)
C3.M <- FindMarkers(filtered_object, ident.1 = 3, min.pct= 0.25, logfc.threshold = 0.25,test.use = "roc",only.pos = TRUE)
C4.M <- FindMarkers(filtered_object, ident.1 = 4, min.pct= 0.25, logfc.threshold = 0.25,test.use = "roc",only.pos = TRUE)
C5.M <- FindMarkers(filtered_object, ident.1 = 5, min.pct= 0.25, logfc.threshold = 0.25,test.use = "roc",only.pos = TRUE)
C6.M <- FindMarkers(filtered_object, ident.1 = 6, min.pct= 0.25, logfc.threshold = 0.25,test.use = "roc",only.pos = TRUE)
C7.M <- FindMarkers(filtered_object, ident.1 = 7, min.pct= 0.25, logfc.threshold = 0.25,test.use = "roc",only.pos = TRUE)
C8.M <- FindMarkers(filtered_object, ident.1 = 8, min.pct= 0.25, logfc.threshold = 0.25,test.use = "roc",only.pos = TRUE)
C9.M <- FindMarkers(filtered_object, ident.1 = 9, min.pct= 0.25, logfc.threshold = 0.25,test.use = "roc",only.pos = TRUE)
C10.M <- FindMarkers(filtered_object, ident.1 = 10, min.pct= 0.25, logfc.threshold = 0.25,test.use = "roc",only.pos = TRUE)
C11.M <- FindMarkers(filtered_object, ident.1 = 11, min.pct= 0.25, logfc.threshold = 0.25,test.use = "roc",only.pos = TRUE)
C12.M <- FindMarkers(filtered_object, ident.1 = 12, min.pct=
0.25,logfc.threshold = 0.25,test.use = "roc",only.pos = TRUE)
C13.M <- FindMarkers(filtered_object, ident.1 = 13, min.pct= 0.25,  logfc.threshold = 0.25,test.use = "roc",only.pos = TRUE)
C14.M <- FindMarkers(filtered_object, ident.1 = 14, min.pct= 0.25,  logfc.threshold = 0.25,test.use = "roc",only.pos = TRUE)
C15.M <- FindMarkers(filtered_object, ident.1 = 15, min.pct= 0.25,  logfc.threshold = 0.25,test.use = "roc",only.pos = TRUE)
C16.M <- FindMarkers(filtered_object, ident.1 = 16, min.pct= 0.25,  logfc.threshold = 0.25,test.use = "roc",only.pos = TRUE)
C17.M <- FindMarkers(filtered_object, ident.1 = 17, min.pct= 0.25,  logfc.threshold = 0.25,test.use = "roc",only.pos = TRUE)
```
# Load rlang
```{r}
library(rlang)
```
# Load tidyverse
```{r}
library(tidyverse)
```
# Top 2 markers
```{r}
top_2_markers <- cluster_markers %>%
  group_by(cluster) %>%
  top_n(n = 2, wt = avg_log2FC)
```
# Extract the gene names
```{r}
top_genes <- top_2_markers$gene
```
# Heatmap for the top2markers
```{r}
DoHeatmap(object = filtered_object, features = top_genes) +
  scale_fill_gradientn(colors = c("blue", "white", "red"))
```
# Violin plot for specific genes:"OGN", "NAALADL2-AS2"
```{r}
VlnPlot(object = filtered_object, 
        features = c("OGN", "NAALADL2-AS2"), 
        group.by = "seurat_clusters", 
        pt.size = 0.1) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
# Violin plot for specific genes:"TMPRSS11E", "CST4"
```{r}
VlnPlot(object = filtered_object, 
        features = c("TMPRSS11E", "CST4"), 
        group.by = "seurat_clusters", 
        pt.size = 0.1) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
# Violin plot for specific genes:"SCN1A", "NTNG1"
```{r}
VlnPlot(object = filtered_object, 
        features = c("SCN1A", "NTNG1"), 
        group.by = "seurat_clusters", 
        pt.size = 0.1) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
# Instpect clusters levels
```{r}
table(filtered_object$seurat_clusters)
head(filtered_object@meta.data)
top_genes <- unique(top_genes)
print(top_genes)
Idents(filtered_object) <- "seurat_clusters"
```

# Dot plot for the top2markers
```{r}
DotPlot(object = filtered_object, features = top_genes, group.by = "seurat_clusters") +
  scale_color_gradientn(colors = c("blue", "white", "red")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
# Features plots: Luminal epithelial cells:KLK3 (PSA), KLK2, FOLH1 (PSMA), ACPP, KRT8, KRT18
```{r}
FeaturePlot(filtered_object, features = c("KLK3 (PSA)", "KLK2", "FOLH1(PSMA)","ACPP", "KRT8","KRT18"))
```
# Violin plot for specific genes
```{r}
VlnPlot(object = filtered_object, 
        features = c("KLK3 (PSA)", "KLK2", "FOLH1(PSMA)","ACPP", "KRT8","KRT18"), 
        group.by = "seurat_clusters", 
        pt.size = 0.1) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
# Feature plot: Basal epithelial cells:KRT5, KRT14, TP63
```{r}
FeaturePlot(filtered_object, features = c("KRT5", "KRT14", "TP63"))
```
# Violin plot for specific genes
```{r}
VlnPlot(object = filtered_object, 
        features = c("KRT5", "KRT14", "TP63"), 
        group.by = "seurat_clusters", 
        pt.size = 0.1) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
# Feature plot: Neuroendocrine  cells:CHGA, CHGB, SYP
```{r}
FeaturePlot(filtered_object, features = c("CHGA", "CHGB", "SYP"))
```
# Violin plot for specific genes
```{r}
VlnPlot(object = filtered_object, 
        features = c("CHGA", "CHGB", "SYP"), 
        group.by = "seurat_clusters", 
        pt.size = 0.1) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
# Feature plot: Stroma cells ( made up of fibroblast, smooth muscles and stromal cell):DCN, LUM, ACTA2
```{r}
FeaturePlot(filtered_object, features = c("DCN", "LUM", "ACTA2"))
```
# Violin plot for specific genes
```{r}
VlnPlot(object = filtered_object, 
        features = c("DCN", "LUM", "ACTA2"), 
        group.by = "seurat_clusters", 
        pt.size = 0.1) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
# Feature plot: immune cells(T Lymphocytes) CD3D, CD3E
```{r}
FeaturePlot(filtered_object, features = c("CD3D", "CD3E"))
```
# Violin plot for specific genes
```{r}
VlnPlot(object = filtered_object, 
        features = c("CD3D", "CD3E"), 
        group.by = "seurat_clusters", 
        pt.size = 0.1) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
# Feature plot: immune cells(B Lymphocytes) CD79A, MS4A1
```{r}
FeaturePlot(filtered_object, features = c("CD79A", "MS4A1"))
```
# Violin plot for specific genes
```{r}
VlnPlot(object = filtered_object, 
        features = c("CD79A", "MS4A1"), 
        group.by = "seurat_clusters", 
        pt.size = 0.1) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
# Feature plot: immune cells(macrophages) CD68, CD163
```{r}
FeaturePlot(filtered_object, features = c("CD68", "CD163"))
```
# Violin plot for specific genes
```{r}
VlnPlot(object = filtered_object, 
        features = c("CD68", "CD163"), 
        group.by = "seurat_clusters", 
        pt.size = 0.1) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
# Feature plot: Mast cells(macrophages) TPSAB1, KIT
```{r}
FeaturePlot(filtered_object, features = c("TPSAB1", "KIT"))
```
# Violin plot for specific genes
```{r}
VlnPlot(object = filtered_object, 
        features = c("TPSAB1", "KIT"), 
        group.by = "seurat_clusters", 
        pt.size = 0.1) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
# Feature plot: Endothelial cells:PECAM1 (CD31), VWF
```{r}
FeaturePlot(filtered_object, features = c("PECAM1 (CD31)", "VWF"))
```
# Violin plot for specific genes
```{r}
VlnPlot(object = filtered_object, 
        features = c("PECAM1 (CD31)", "VWF"), 
        group.by = "seurat_clusters", 
        pt.size = 0.1) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
# Feature plot: Luminal progenitor cells:TACSTD2 (TROP2), KRT4, PSCA
```{r}
FeaturePlot(filtered_object, features = c("TACSTD2 (TROP2)", "KRT4","PSCA"))
```
# Violin plot for specific genes
```{r}
VlnPlot(object = filtered_object, 
        features = c("PECAM1 (CD31)", "VWF"), 
        group.by = "seurat_clusters", 
        pt.size = 0.1) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
# Step to confirm the assigned clusters
```{r}
FeaturePlot(filtered_object, features = c("KLK3", "KRT5", "CHGA", "DCN", "CD3D"))
```
# Average expression per cluster
```{r}
AverageExpression(filtered_object, features = c("KLK3", "KRT5", "CHGA", "DCN", "CD3D"), return.seurat = TRUE)
```
# Manually assign cluster names
# Modify new cluster id to include 18 cell types names coreesponding to the number of clusters
```{r}
new.cluster.ids <- c("Luminal", "Basal", "Neuroendocrine", "Stroma", "T cells", "B cells", 
  "Macrophages", "Mast cells", "Endothelial", "Luminal Progenitor",
  "Luminal", "Basal", "Neuroendocrine", "Stroma", "T cells", "B cells", 
  "Macrophages", "Endothelial")
```
# Verify no of clusters
```{r}
length(levels(filtered_object))
```
# Assign names to clusters
```{r}
names(new.cluster.ids) <- levels(filtered_object)
```
# rename clusters
```{r}
filtered_object <- RenameIdents(filtered_object, new.cluster.ids)
```
# Confirming that renaming worked
```{r}
table(Idents(filtered_object))
```
# visualize the renamed clusters
```{r}
DimPlot(filtered_object, reduction = "umap", label = TRUE, pt.size = 0.5) + 
  ggtitle("UMAP Visualization of Renamed Clusters")
```
# Dot plot general vs cell types
```{r}
DotPlot(object = filtered_object, features = top_genes) +
  scale_color_gradientn(colors = c("blue", "white", "red")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
# Get differentially expressed markers per cell type
```{r}
cluster_markers <- FindMarkers(filtered_object, ident.1 = "Luminal Progenitor", ident.2 = "Endothelial", ident.3 = "Mast cells", ident.4 = "Macrophages", ident.5 = "B cells", ident.6 = "T cells", ident.7 = "Stroma", ident.8 = "Neuroendocrine", ident.9 = "Basal", ident.10 = "Luminal")
```
# Load ClusterProfiler library
```{r}
library(clusterProfiler)
```
# Load org.Hs.eg.db 
```{r}
library(org.Hs.eg.db)
```
# Extract top-markers
```{r}
Extract_Top_Markers <- function(marker_dataframe, num_genes, group_by, rank_by, gene_column) {
  marker_dataframe %>%
    dplyr::group_by(across(all_of(group_by))) %>% 
    dplyr::arrange(desc(across(all_of(rank_by)))) %>% 
    dplyr::slice_head(n = num_genes) %>% 
    dplyr::ungroup() %>% 
    dplyr::select(all_of(gene_column), everything())
}
```
# Is extract_top_marker in dataframe?
```{r}
class(Extract_Top_Markers)
```
# Assign FindAllMarker output to a dataframe
```{r}
Extract_Top_Markers <- FindAllMarkers(
  object = filtered_object,
  only.pos = TRUE,
  min.pct = 0.25,
  logfc.threshold = 0.25
)
```
#check if cluster_markers is in dataframe
```{r}
class(Extract_Top_Markers)
```
# Convert row names to a column named 'gene'
```{r}
cluster_markers$gene <- rownames(cluster_markers)
```
# Confirm the change
```{r}
head(cluster_markers$gene)
```
# Check how many genes in cluster_markers$gene match raw_object
```{r}
sum(cluster_markers$gene %in% rownames(filtered_object))
```
# View unmatched genes
```{r}
unmatched_genes <- cluster_markers$gene[!cluster_markers$gene %in% rownames(filtered_object)]
head(unmatched_genes)
```
# Check the identities
```{r}
head(Idents(filtered_object))
table(Idents(filtered_object))
```
# Add cell_type column using a mapping approach
```{r}
cluster_markers$cell_type <- Idents(filtered_object)[cluster_markers$gene]
```
# View a sample of the data
```{r}
head(cluster_markers)
```
# Match genes in cluster_markers with raw_object rownames
```{r}
cluster_markers$cell_type <- filtered_object@meta.data$cell_type[match(cluster_markers$gene, rownames(filtered_object))]
```
# Filter markers by statistical significance
```{r}
significant_markers <- cluster_markers %>%
  filter(p_val_adj < 0.05)
```
# Use fold change to select markers with biologically meaningful expression
```{r}
selected_markers <- significant_markers %>% 
  filter(avg_log2FC > 0.5)
```



